Quantitative Benchmarks for Theory Exploration

### 2017-04-24 PhDSymposium2017/abstract/abstract.tex
Symposium updates

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Quantitative Benchmarks for Theory Exploration},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}
\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{"#1"}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

@@

\begin{document}
\maketitle
\begin{abstract}
  \emph{Automated theory exploration} has the potential to help scientists,
  mathematicians and programmers discover unexpected properties of their models,
  theories and programs; but has so far been limited to small-scale experiments,
  with hand-picked data and qualitative results. We address these problems by
  introducing a novel benchmark suite, orders of magnitude larger than previous
  data sets, and use it to evaluate a variety of theory exploration approaches
  (both existing and novel) and measure how they scale. It is our hope that the
  availability of a robust method for comparing systems will encourage
  competition and innovation in this field, as it has done for e.g.\ automated
  theorem proving.
\end{abstract}

@@

Many fields represent knowledge using \emph{formal systems}, for example
scientific models, mathematical theories and computer programs. As requirements
grow, these systems can be modified in two ways: becoming more complicated, and
becoming more numerous. Complication makes a system's behaviour harder to
predict, whilst proliferation dilutes the attention that can be given to each
particular system. This impacts aspects like safety and security, for example
proving that a program will never leak confidential information, or that a model
of some mechanism will never enter an unsafe configuration.

One way to mitigate these problems is through automation: \emph{automated theory
exploration} is the task of discovering interesting properties and behaviours of
a given formal system, such as a software library or set of equations. Such
tools can be used by domain experts, to discover structure, redundancy and
patterns in a system; and, more importantly, to be \emph{surprised} by
unexpected insights.

@@

\begin{figure}
  \begin{equation*}
    plus(zero,    y) = y \\
    plus(succ(x), y) = succ(plus(x, y)) \\
    length(nil)        = zero \\
    length(cons(x, y)) = succ(length(y)) \\
    append(nil,       z) = z \\
    append(cons(x,y), z) = cons(x, append(y, z)) \\
    map(f, nil)        = nil \\
    map(f, cons(x, y)) = cons(f(x), map(f, y))
  \end{equation*}
  \caption{Functions.}
  \label{fig:defs}
\end{figure}

Our running example will be the software library\footnote{Our treatment of
  software is directly applicable to mathematics \cite{wadler2015propositions},
  and is also applicable to scientific models if we encode them as programs
  which output predicted observations \cite{davies1990physical}
} in figure~\ref{fig:defs}

functions for manipulating data. This

We can think of this as an axiomatic equational theory; it's also We will concentrate on the example of a software library, but this is

The problem of \emph{automated theorem proving} (ATP) was one of the first tasks
studied in the field of artificial intelligence \cite{newell1956logic,
  sutcliffe2001evaluating}: given some statement about a formal system, prove
that it is true. Many ATP approaches have been devised, and systems are compared
by evaluating their performance on large benchmark problems, such as TPTP
(Thousands of Problems for Theorem Provers).

In contrast, theory \emph{exploration}progress in ATP has been through progTheory theorem proving a the  experts with a system for  discovering the behaviour of such systems has impact may  can exhibit Both present challenges to understandingare As these systems become more complexcomputers become ever more important in research, industry and
society, so does the ability to program them effectively, economically
and \emph{correctly}. Whilst tools exist to help programmers understand
and manipulate software, ranging from linters \cite{Johnson78lint} through to
formal verification systems \cite{boyer1983proof}, these are often limited in
scope (e.g.\ to syntax rather than semantics) or require precise guidance from
the user. One reason is that automating the search for patches
\cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings, optimisations
\cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other
routine programming tasks is prohibitively expensive due to the size of the
search space.

We investigate the use of machine learning techniques to restrict such
search spaces more intelligently than uninformed local search,
specifically the use of \emph{clustering} to aid \emph{theory
exploration} (TE) in discovering program equivalences.

@@

Our work builds on the existing TE system
\textsc{QuickSpec}\cite{QuickSpec}, which finds equivalent expressions
in the Haskell programming language \cite{marlow2010haskell}, built from a user-provided
\emph{signature} of constants and variables. For example, given a
signature of the list-manipulating functions \texttt{reverse},
\texttt{length} and \texttt{append}, plus two list variables \texttt{x}
and \texttt{y}, \textsc{QuickSpec} will discover that
\texttt{reverse\ (reverse\ x)} is equivalent to \texttt{x}; that
\texttt{length\ (reverse\ x)} is equivalent to \texttt{length\ x}; that
\texttt{length\ (append\ x\ y)} is equivalent to
\texttt{length\ (append\ y\ x)}; and so on. These are discovered by
enumerating all well-typed combinations of the signature's constants and
variables, then randomly instantiating the variables \cite{claessen2011quickcheck} and comparing the
resulting closed terms. Any expressions which remain indistiguishable
after hundreds of instantiations are conjectured to be equivalent and,
after removing redundancies, are produced as output. These conjectures
can be sent through a theorem prover \cite{claessen2013automating} and used to inform tests,
optimisations, refactoring, verification, or simply to help the
programmer learn more about the code.

@@

To reduce the user's need to cherry-pick signatures, our approach
accepts a large signature (e.g.\ a complete Haskell package) and uses
statistical machine learning methods to select smaller signatures
automatically, using a \emph{recurrent clustering} method inspired by
those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.

@@

These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices} \cite{de1972lambda}). To replace global variables, $\phi$ first performs another round of clustering (hence the name
``recurrent''), without including the current expression; each global
variable is replaced by its cluster index found by this ``inner'' round
of clustering.

@@

\begin{equation*}
  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots
\end{equation*}

@@

We obtain ASTs from Haskell projects using a bespoke plugin for the GHC
compiler. From these ASTs, we extract type information which is used to append variables to the signature; and dependency information, which is used to topologically sort the clustering rounds. We then invoke our iterative algorithm, which alternates between feature extraction and clustering until all elements of the signature have been clustered. Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which a) have an argument type which we can randomly generate and b) have an output type which we can compare. \textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.

>>>>

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{enumitem}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum0\ifxetex{1}\fi\ifluatex{1}\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex{
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  }\fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex%
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Quantitative Benchmarks for Theory Exploration},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

\newcommand{\name}[1]{\mathrm{#1}}

\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{``#1''}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

\newlist{inlinelist}{enumerate*}{1}
\setlist*[inlinelist,1]{%
  label= (\roman*),
}

@@

\begin{document}
\maketitle
\begin{abstract}
  \emph{Automated theory exploration} has the potential to help scientists,
  mathematicians and programmers discover unexpected properties of their models,
  theories and programs; but has so far been limited to small-scale experiments,
  with hand-picked data and qualitative results. We address these problems by
  introducing a novel benchmark suite, orders of magnitude larger than previous
  data sets, and use it to evaluate a variety of theory exploration approaches
  (both existing and novel) and measure how they scale. It is our hope that the
  availability of a robust method for comparing systems will encourage
  competition and innovation in this field.
\end{abstract}

@@

Many fields represent knowledge using \emph{formal systems}, for example
scientific models, mathematical theories and computer programs. As requirements
increase, these systems can grow in two ways: becoming more complex, and
becoming more numerous. Complexification makes a system's behaviour harder to
predict, whilst proliferation dilutes the attention that can be given to each
particular system. This impacts aspects like safety and security, for example
proving that software will not leak confidential information, or that mechanisms
will not enter unsafe configurations.

One way to handle this growth is through automation: formal systems are easily
represented in software, and techniques like \emph{automated theory exploration}
(ATE) and \emph{automated theorem proving} (ATP) can help experts to discovery
and verify the properties they do or do not want the system to exhibit. ATP is a
mature field, with many competing implementations and competitive benchmarks. In
contrast, ATE has seen little attention, and existing systems have mostly been
evaluated qualitatively, on only a few hand-picked examples.

We propose a benchmarking methodology for ATE systems, to evaluate and compare
their performance on many large problems. We apply our benchmark to existing ATE
systems, as well as two novel approaches.

@@

\begin{figure}
  \begin{equation*}
    \begin{aligned}
      \name{plus}(\name{zero},         y) &= y                                   \\
      \name{plus}(\name{succ}(x),      y) &= \name{succ}(\name{plus}(x, y))      \\
      \name{length}(\name{nil})           &= \name{zero}                         \\
      \name{length}(\name{cons}(x, y))    &= \name{succ}(\name{length}(y))       \\
      \name{append}(\name{nil},        z) &= z                                   \\
      \name{append}(\name{cons}(x,y),  z) &= \name{cons}(x, \name{append}(y, z)) \\
      \name{map}(f,           \name{nil}) &= \name{nil}                          \\
      \name{map}(f,    \name{cons}(x, y)) &= \name{cons}(f(x), \name{map}(f, y))
    \end{aligned}
  \end{equation*}
  \caption{Definitions for $\name{plus}$, $\name{length}$, $\name{append}$ and
    $\name{map}$. $\name{zero}$ and $\name{succ}$, $\name{nil}$ and
    $\name{cons}$ are symbols encoding natural numbers and lists, respectively.
    $\name{constants}$ and $variables$ are distinguished by
    typeface.}\label{fig:defs}
\end{figure}

Our running example will be the small software library defined in
figure~\ref{fig:defs}\footnote{Our approach can also be applied directly to
  mathematical theories~\cite{wadler2015propositions}, and to scientific models
  encoded as programs~\cite{davies1990physical}}. One property of this library
is that the following programs are equivalent (for all $f$, $x$ and $y$):

\begin{equation}\label{eq:mapappend}
  \name{map}(f, \name{append}(x, y)) =
    \name{append}(\name{map}(f, x), \name{map}(f, y))
\end{equation}

The $\name{map}$ function applies an arbitrary function to all of the elements
in a list; $\name{append}$ joins one list to the end of another. Property
\ref{eq:mapappend} states that $\name{map}$ distributes over $\name{append}$,
i.e.\ we can perform $\name{map}$ and $\name{append}$ in either order and get
the same result. This is useful, since it may be much faster to run
$\name{map}(f, x)$ and $\name{map}(f, y)$ in parallel on different machines,
then $\name{append}$ the results; compared to $\name{append}$ing the inputs and
running one big $\name{map}$.\footnote{This is one half of the ``MapReduce''
  model for parallel programming.~\cite{dean2008mapreduce}}

There are two aspects to finding these properties:

\begin{enumerate}
\item We must \emph{state} the property, singling it out for attention, because
  it is somehow \emph{interesting}.

\item We must \emph{demonstrate} the property, providing evidence such as a
  formal proof or empirical tests.
\end{enumerate}

\subsection{Automated Theorem Proving}

Much work has been done on \emph{demonstrating} such properties, through the use
of automated theorem proving. ATP has been an active area of study in artificial
intelligence since the field was founded~\cite{newell1956logic,
  sutcliffe2001evaluating}, and modern ATP systems such as CVC4, SPASS and
Vampire take part in regular competitions~\cite{CASC}: attempting to prove as
much as possible, as quickly as possible, from large benchmark suites such as
TPTP (Thousands of Problems for Theorem Provers)~\cite{TPTP}.

\subsection{Theory Exploration}

We focus on the less-developed problem of \emph{conjecturing} properties:
choosing some particular property, from the infinite number of possibilities,
because it is somehow ``interesting''. Many measures of interestingness have
been proposed\cite{INTERESTING}; John Conway proposed that a good conjecture
should be ``outrageous''\cite{CONWAY}, but we also need some degree of
\emph{plausibility} to avoid obvious falsehoods, even if we defer full proofs
to a later stage.

A particularly simple way to check if a property is interesting is to see if a
human has already bothered to state it. If the theory given to an ATE system has
already been studied by mathematicians, we can judge the quality of the system's
output by comparing it against a corpus of known properties.

Existing ATE systems have taken this approach, using theories and properties
from the Isabelle proof assistant. Unfortunately, only two theories (natural
numbers and lists) have been used to compare systems, and these only contain
FIXME properties. This is not enough data to draw meaningful comparisons or
extrapolations from.

\section{Benchmarking}

We propose a larger benchmark suite, containing many more definitions and
properties, similar to those which drive competition in ATP. In fact, we can
take the definitions and theorems from existing ATP benchmarks and repurpose
them as the corpus of an ATE benchmark! We have chosen the TIP (Tons of
Inductive Problems) benchmark~\cite{TIP} as the corpus for our approach, since
it \begin{inlinelist}
\item separates definitions from theorems
\item allows higher-order functions (like $\name{map}$)
\item incorporates many tricky problems from the theorem-proving literature
\end{inlinelist}.

Each of TIP's FIXME problems contains a set of definitions and a single theorem.
We collect up all of the distinct definitions from all of the problems, to form
one large theory. We collect up the theorems as our corpus, to compare ATE
results against. Since FIXME is far larger than previously attempted theories,
we also provide tools to uniformly sample a subset of definitions and their
theorems.

\section{Results}

Figure FIXME compares the median run times of the existing
\textsc{QuickSpec}\cite{QuickSpec} system across a variety of theory sizes,
sampled from our benchmark. Figure FIXME shows the average proportion of corpus
properties that were found for each sample. sizesFigure FIXME shows the run times of various
\section{Background}\label{background}

@@

Our work builds on the existing TE system
, which finds equivalent expressions
in the Haskell programming language~\cite{marlow2010haskell}, built from a user-provided
\emph{signature} of constants and variables. For example, given a
signature of the list-manipulating functions \texttt{reverse},
\texttt{length} and \texttt{append}, plus two list variables \texttt{x}
and \texttt{y}, \textsc{QuickSpec} will discover that
\texttt{reverse\ (reverse\ x)} is equivalent to \texttt{x}; that
\texttt{length\ (reverse\ x)} is equivalent to \texttt{length\ x}; that
\texttt{length\ (append\ x\ y)} is equivalent to
\texttt{length\ (append\ y\ x)}; and so on. These are discovered by
enumerating all well-typed combinations of the signature's constants and
variables, then randomly instantiating the variables~\cite{claessen2011quickcheck} and comparing the
resulting closed terms. Any expressions which remain indistiguishable
after hundreds of instantiations are conjectured to be equivalent and,
after removing redundancies, are produced as output. These conjectures
can be sent through a theorem prover~\cite{claessen2013automating} and used to inform tests,
optimisations, refactoring, verification, or simply to help the
programmer learn more about the code.

@@

To reduce the user's need to cherry-pick signatures, our approach
accepts a large signature (e.g. a complete Haskell package) and uses
statistical machine learning methods to select smaller signatures
automatically, using a \emph{recurrent clustering} method inspired by
those of \textsc{ML4PG}~\cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)}~\cite{Heras.Komendantskaya.Johansson.ea:2013}.

@@

These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices}~\cite{de1972lambda}). To replace global variables, $\phi$ first performs another round of clustering (hence the name
``recurrent''), without including the current expression; each global
variable is replaced by its cluster index found by this ``inner'' round
of clustering.

@@

\begin{equation*}
  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots )
\end{equation*}

@@

We obtain ASTs from Haskell projects using a bespoke plugin for the GHC
compiler. From these ASTs, we extract type information which is used to append variables to the signature; and dependency information, which is used to topologically sort the clustering rounds. We then invoke our iterative algorithm, which alternates between feature extraction and clustering until all elements of the signature have been clustered. Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which
\begin{inlinelist}
\item have an argument type which we can randomly generate
\item have an output type which we can compare
\end{inlinelist}. \textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.
